An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models

### 2025-05-02 Project_0/manuscript/manuscript.Rmd
project 0 draft to submit

---
title: An investigation of factors affecting inferences from and reliablity of state-space age-structured assessment models
author: 
  - Timothy J. Miller^1,2^
  - Greg Britten^3^
  - Elizabeth N. Brooks^2^
  - Gavin Fay^4^
  - Alex Hansell^2^
  - Christopher M. Legault^2^
  - Chengxue Li^2^
  - Brandon Muffley^5^
  - Brian C. Stock^6^
  - John Wiedenmann^7^
date: "`r format(Sys.time(), '%d %B, %Y')`"
fontsize: 12pt
output:
  pdf_document:
#    citation_package: natbib
    fig_caption: yes
    includes:
      in_header:
      - options.sty
      - journalshortcuts.tex
    keep_tex: yes
    number_sections: yes
  html_document:
    df_print: paged
#  word_document: default
header-includes:
#- \usepackage{makecell}
csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
bibliography: manuscript.bib
#biblio-style: cjfas2.bst
---

@@

State-space models are increasingly used for stock assessment, and evaluations of their statistical reliability and best practices for selecting among process error configurations are needed. We simulated 72 operating models that varied fishing pressure and observation error across process errors in recruitment, survival, selectivity, catchability, and/or natural mortality. We fit estimating models with different assumptions on the process error source and whether median natural mortality or a stock-recruit relationship were estimated. Estimating models without a stock-recruit relationship that assumed the correct process error source and median natural mortality had high convergence rates and low bias. Bias was also low under many incorrect process error assumptions when there was contrast in fishing pressure and low observation error. Marginal AIC most accurately distinguished process errors on recruitment, survival, and selectivity, as well as larger magnitude process errors of other types. Retrospective patterns were generally small but were sizable for recruitment when observation error was high. These results help establish the statistical reliability of state space assessment models and pave the way for the next-generation of fisheries stock assessment

@@

For a given model attribute we calculated the relative error 
$$
\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
$$
from fitting a given estimating model to simulated data set $i$ configured for a given OM where $\widehat \theta_i$ and $\theta_i$ are the estimated and true values for simulation $i$. We estimated bias as the median of the relative errors across all simulations for a given OM and EM combination. We constructed 95\% confidence intervals for the median relative bias using the binomial distribution approach as in @millerhyun18 and @stockmiller21. We present results for bias in terminal year estimates of SSB and recruitment, Beverton-Holt stock recruit parameters ($a$ and $b$), and median natural mortality rate. Results for terminal year fishing mortality are were strongly negatively correlated with those for SSB and are provided in the Supplementary Materials.

@@

For R and R+S OMs, convergence generally declined for most EMs when the median natural mortality rate was estimated and/or the Beverton-Holt SRR was estimated even when the process error assumptions of the EMs and OMs matched (Figure \ref{hessian_SE_convergence}, A). When there was high observation error and constant fishing pressure ($F = F_{\text{MSY}}$ for all 40 years), convergence was poor for of all EM process error configurations other than R EMs when fitted to R OMs ($\sigma_{2+} = 0$) regardless of whether median natural mortality and SRRs were estimated. Convergence of R EMs was high for all R and R+S OMs except when there was high observation error and constant fishing pressure, and when median natural mortality and SRRs were estimated. R+S EMs fit to R OMs exhibited poor convergence regardless of whether natural mortality or a SRR was estimated. R+S EMs fit to R+S OMs had highest convergence rates when there was contrast in fishing pressure and low observation error. Convergence rates were high for all EMs when fit to data from R+S OMs with lower observation error except those where median natural mortality and/or SRRs were estimated.

@@

Across all OMs, there was generally better accuracy (less bias and/or lower variability) for estimation of the Beverton Holt $a$ parameter than the $b$ parameter. In R and R+S OMs, EMs with the correct assumptions about process errors provided the least biased estimation of Beverton-Holt SRR parameters when there was a change in fishing pressure over time and lower variability of recruitment process errors, but there was little effect of estimating median natural mortality and a small increase in bias for those OMs that had high observation error (Figure \ref{SR_rel_error}, A). For other R and R+S OMs, estimating natural mortality often resulted in less biased estimation of SRR parameters. There was generally large variability in relative errors of the SRR parameter estimates, but the lowest variability occurred with low variability in recruitment and little or no variability in survival process errors ($\sigma_{2+} \in \{0,0.25\}$), and contrast in fishing pressure.

@@

Bias for SRR parameters was large and variability in relative errors was greatest for most EMs fit to R+Sel OMs with constant fishing pressure (Figure \ref{SR_rel_error}, C). Less bias in parameter estimation occurred for OMs with a change in fishing pressure and the best accuracy occurred for those OMs that had low observation error and more variable and uncorrelated selectivity process errors, and when the EMs hadd with the correct process error assumption. There was little effect of estimating natural mortality on relative errors for SRR parameters.

@@

Like R+Sel OMs, all EM process error types produced accurate estimation of median natural mortality rate when fit to R+q OMs with contrast in fishing pressure, low observation error and low catchability process error variability (Figure \ref{M_rel_error}, D). Most EM process error types produced biased estimation of median natural mortality when R+q OMs had high observation error and constant fishing pressure.

@@

Often poor convergence result when parameter estimates are at their bounds [@carvalhoetal21], and this also applies to variance parameters for random effects with state-space assessment models. Even when Hessian is invertible, parameters that are poorly informed will have extremely large variance estimates. This further inspection can lead to a more appropriate and often more parsimonious model configuration where the problematic parameters are not estimated. For example, process error variance parameters that are estimated close to 0 indicates that the random effects are estimated to have little or no variability and removing these process errors is warranted. Generally, our results suggest we can expect lower probability of convergence of state-space assessment models when estimating natural mortality or SRRs because of the difficulty distinguishing these parameters from others being estimated in assessment model with data that are typically available. Our experiments did not aim to emulate the practitioner decision process in developing model configurations (e.g. removing a source of process error and refitting the model when process error variance parameters were estimated close to 0). Evaluating the efficacy of such a decision process when applying Ems might be important in closed loop simulations (e.g. MSE) aimed at quantifying management performance.

A factor affecting the convergence criteria, particularly for maximum likelihood estimation of models with random effects, is numerical accuracy. All optimizations performed in these simulations are of the Laplace approximation of the marginal likelihood and, therefore, gradients and Hessians are also with respect to this approximation (see TMB::sdreport in the Template Model Builder package). Functionality within the Template Model Builder package exists (i.e., TMB::checkConsistency) to check the validity of the Laplace approximation and the utility of this as a diagnostic for state-space assessment models should be explored further. Furthermore, numerical methods are used to calculate and invert the Hessian for variance estimation for models with random effects. Along with our results, the potential lack of accuracy imposed by these approximations, suggests at least investigating whether the Hessian is positive definite when the calculated absolute gradients are not terribly large.

@@

Of the OM process error configurations we considered, we found AIC to be accurate for selecting models with process errors on recruitment and apparent survival (R and R+S). Fitting models to other OMs rarely preferred R+S EMs, and R and R+S EMs were nearly always selected for the matching OMs; a similar result was reported by @liljestrandetal24. For other sources of process error, accuracy of AIC was improved to useful levels when there was larger variability in the process errors and/or lower observation error.

@@

Although we did not compare models with alternative SRRs (e.g., Ricker and Beverton-Holt), we do not expect AIC to perform any better distinguishing between relationships. Our finding that AIC tended to choose simpler recruitment models in most cases contrasts with the noted bias in AIC for more complex models [@shibata76;@katz81;@kassraftery95], but, whereas those findings apply to more the much more common comparison of models that are fit to raw and independent observations, here we are comparing state-space models which account for observation error and estimate process errors in latent variables.

@@

As expected, bias in all parameters and assessment output was generally improved with lower observation error. Estimation of SRR parameters was reliable in ideal scenarios of low observation error and contrast in fishing for some R+Sel and R+M OMs, but generally estimation was biased and(or) highly variable. We found substantial bias in estimated SRR parameters in R and R+S OMs particularly with high variability in recruitment and apparent survival process errors, suggesting that practitioners should be cautious when models without the SRR suggest this to be the case.

On the other hand, estimation of median natural mortality was reliable in many OM scenarios with contrast in fishing pressure and in some OMs EMs that also estimated the SRR parameters bias for those parameters was improved.  Conversely, for some R+Sel and R+q OMs where there was bias in natural mortality due to high observation error, estimating the SRR reduced the bias in median natural mortality rate.  However, estimating median natural mortality did cause poor accuracy in SSB estimation in many OMs without contrast in fishing pressure over time and with higher observation error. Thus, estimating median natural mortality should be approached with caution in state-space assessment models, particularly given its significant impact on determination of reference point and stock status [@lietal24].

@@

Our simulation study examined the importance of several factors for reliable inferences from state-space age-structured assessment models. Across all factors, AIC accurately distinguished models with process errors on recruitment only or on recruitment and apparent survival. Accuracy for other process error types required a strong signal (high process variability) with low noise (low observation uncertainty). Therefore, we expect practitioners will find R+S configurations to provide satisfactory diagnostics across a range of life history and data quality scenarios. Contrast in fishing pressure was consistently an important factor across all measures of reliability we examined. AIC generally performed poorly for selecting the SRR but performance was improved with low recruitment variability and contrast in fishing pressure. However, some bias in estimation in at least one of the SRR parameters existed in nearly all OM-EM combinations. Because bias in terminal SSB and retrospective patterns were indifferent to whether or not the SRR was estimated, and convergence was slightly better without the SRR, a sensible default would be to fit models without an assumed SRR.

@@

\begin{landscape}
\begin{figure}
\begin{center}
\includegraphics{term_SSB_bias_plots}
\end{center}
\caption{Median relative error of terminal year SSB for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D).  Diamond shaped points denote results where the EM process error assumption matches that of the operating model. Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SSB_rel_error}
\end{figure}
\end{landscape}

>>>>

---
title: An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models
author: 
  - Timothy J. Miller^1,2^
  - Greg Britten^3^
  - Elizabeth N. Brooks^2^
  - Gavin Fay^4^
  - Alex Hansell^2^
  - Christopher M. Legault^2^
  - Chengxue Li^2^
  - Brandon Muffley^5^
  - Brian C. Stock^6^
  - John Wiedenmann^7^
date: "`r format(Sys.time(), '%d %B, %Y')`"
fontsize: 12pt
output:
  pdf_document:
#    citation_package: natbib
    fig_caption: yes
    includes:
      in_header:
      - options.sty
      - journalshortcuts.tex
    keep_tex: yes
    number_sections: yes
  html_document:
    df_print: paged
#  word_document: default
header-includes:
#- \usepackage{makecell}
csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
bibliography: manuscript.bib
#biblio-style: cjfas2.bst
---

@@

State-space models are increasingly used for stock assessment, and evaluations of their statistical reliability and best practices for selecting among process error configurations are needed. We simulated 72 operating models that varied fishing pressure and observation error across process errors in recruitment, survival, selectivity, catchability, and/or natural mortality. We fit estimating models with different assumptions on the process error source and whether median natural mortality or a stock-recruit relationship were estimated. Estimating models without a stock-recruit relationship that assumed the correct process error source and median natural mortality had high convergence rates and low bias. Bias was also low under many incorrect process error assumptions when there was contrast in fishing pressure and low observation error. Marginal AIC most accurately distinguished process errors on recruitment, survival, and selectivity, and other process error sources when variability was greater. Retrospective patterns were generally small but were sizable for recruitment when observation error was high. These results help establish the statistical reliability of state space assessment models and pave the way for the next-generation of fisheries stock assessment.

@@

For a given model attribute we calculated the relative error 
$$
\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
$$
from fitting a given estimating model to simulated data set $i$ configured for a given OM where $\widehat \theta_i$ and $\theta_i$ are the estimated and true values for simulation $i$. We estimated bias as the median of the relative errors across all simulations for a given OM and EM combination. We constructed 95\% confidence intervals for the median relative bias using the binomial distribution approach as in @millerhyun18 and @stockmiller21. We present results for bias in terminal year estimates of SSB and recruitment, Beverton-Holt stock recruit parameters ($a$ and $b$), and median natural mortality rate. Results for terminal year fishing mortality were strongly negatively correlated with those for SSB and are provided in the Supplementary Materials.

@@

For many R and R+S OMs, convergence rate declined when either the median natural mortality rate or the Beverton-Holt SRR was estimated even when the process error assumptions of the EMs and OMs matched (Figure \ref{hessian_SE_convergence}, A). When there was high observation error and constant fishing pressure ($F = F_{\text{MSY}}$ for all 40 years), convergence was poor for of all EM process error configurations other than R EMs when fitted to R OMs ($\sigma_{2+} = 0$) regardless of whether median natural mortality and SRRs were estimated. Convergence of R EMs was high for all R and R+S OMs except when there was high observation error and constant fishing pressure, and when median natural mortality and SRRs were estimated. R+S EMs fit to R OMs exhibited poor convergence regardless of whether natural mortality or a SRR was estimated. R+S EMs fit to R+S OMs had highest convergence rates when there was contrast in fishing pressure and low observation error. Convergence rates were high for all EMs when fit to data from R+S OMs with lower observation error except those where median natural mortality and/or SRRs were estimated.

@@

Across all OMs, there was generally less bias and(or) lower variability in estimation of the Beverton Holt $a$ parameter than the $b$ parameter. In R and R+S OMs, EMs with the correct assumptions about process errors provided the least biased estimation of Beverton-Holt SRR parameters when there was a change in fishing pressure over time and lower variability of recruitment process errors, but there was little effect of estimating median natural mortality and a small increase in bias for those OMs that had high observation error (Figure \ref{SR_rel_error}, A). For other R and R+S OMs, estimating natural mortality often resulted in less biased estimation of SRR parameters. There was generally large variability in relative errors of the SRR parameter estimates, but the lowest variability occurred with low variability in recruitment and little or no variability in survival process errors ($\sigma_{2+} \in \{0,0.25\}$), and contrast in fishing pressure.

@@

Bias for SRR parameters was large and variability in relative errors was greatest for most EMs fit to R+Sel OMs with constant fishing pressure (Figure \ref{SR_rel_error}, C). Less bias in parameter estimation occurred for OMs with a change in fishing pressure and the best accuracy occurred for those OMs that had low observation error and more variable and uncorrelated selectivity process errors, and when the EMs had with the correct process error assumption. There was little effect of estimating natural mortality on relative errors for SRR parameters.

@@

@@

Often poor convergence result when parameter estimates are at their bounds [@carvalhoetal21], and this also applies to variance parameters for random effects with state-space assessment models. Even when the Hessian is invertible, parameters that are poorly informed will have extremely large variance estimates. This further inspection can lead to a more appropriate and often more parsimonious model configuration where the problematic parameters are not estimated. For example, process error variance parameters that are estimated close to 0 indicates that the random effects are estimated to have little or no variability and removing these process errors is warranted. Generally, our results suggest we can expect lower probability of convergence of state-space assessment models when estimating natural mortality or SRRs because of the difficulty distinguishing these parameters from others being estimated in assessment model with data that are typically available. Our experiments did not aim to emulate the practitioner decision process in developing model configurations (e.g. removing a source of process error and refitting the model when process error variance parameters were estimated close to 0). Evaluating the efficacy of such a decision process when applying EMs might be important in closed loop simulations (e.g. MSE) aimed at quantifying management performance.

A factor affecting the convergence criteria, particularly for maximum likelihood estimation of models with random effects, is numerical accuracy. All optimizations performed in these simulations are of the Laplace approximation of the marginal likelihood and, therefore, gradients and Hessians are also with respect to this approximation (see TMB::sdreport in the Template Model Builder package). Functionality within the Template Model Builder package exists (i.e., TMB::checkConsistency) to check the validity of the Laplace approximation and the utility of this as a diagnostic for state-space assessment models should be explored further. Furthermore, numerical methods are used to calculate and invert the Hessian for variance estimation for models with random effects. Along with our results, the potential lack of accuracy imposed by these approximations, suggests at least investigating whether the Hessian is positive definite when the calculated absolute gradients are not terribly large (e.g, $< 1$).

@@

Of the OM process error configurations we considered, we found AIC to be accurate for selecting models with process errors on recruitment and apparent survival (R and R+S). Fitting models to other OMs rarely preferred R+S EMs, and R and R+S EMs were nearly always selected for the matching OMs; a similar result was reported by @liljestrandetal24. For other sources of process error, accuracy of AIC was improved when there was larger variability in the process errors and/or lower observation error.

@@

Although we did not compare models with alternative SRRs (e.g., Ricker and Beverton-Holt), we do not expect AIC to perform any better distinguishing between relationships. Our finding that AIC tended to choose simpler recruitment models in most cases contrasts with the noted bias in AIC for more complex models [@shibata76;@katz81;@kassraftery95], but, whereas those findings apply to the much more common comparison of models that are fit to raw and independent observations, here we are comparing state-space models which account for observation error and estimate process errors in latent variables.

@@

As expected, bias in all parameters and assessment output was generally improved with lower observation error. Estimation of SRR parameters was reliable in ideal scenarios of low observation error and contrast in fishing for some R+Sel and R+M OMs, but generally estimation was biased and(or) highly variable. We found substantial bias in estimated SRR parameters in R and R+S OMs particularly with high variability in recruitment and apparent survival process errors, suggesting that practitioners should be cautious when fitted assessment models have these properties.

On the other hand, estimation of median natural mortality was reliable in many OM scenarios with contrast in fishing pressure, consistent with @hoenigetal25. In some OMs, when EMs estimated the SRR parameters and median natural mortality, bias for those parameters was improved.  Conversely, for some R+Sel and R+q OMs where there was bias in natural mortality due to high observation error, estimating the SRR reduced the bias in median natural mortality rate.  However, estimating median natural mortality did cause poor accuracy in SSB estimation in many OMs without contrast in fishing pressure over time and with higher observation error. Thus, estimating median natural mortality should be approached with caution in state-space assessment models, particularly given its significant impact on determination of reference point and stock status [@lietal24].

@@

Our simulation study examined the importance of several factors for reliable inferences from state-space age-structured assessment models. Contrast in fishing pressure was consistently an important factor across all measures of reliability we examined. AIC accurately distinguished models with process errors on recruitment only (R) or on recruitment and apparent survival (R+S). Accuracy for other process error types required a strong signal (high process variability) with low noise (low observation uncertainty). Therefore, we expect practitioners will find R+S configurations to provide satisfactory diagnostics across a range of life history and data quality scenarios.  AIC generally performed poorly for selecting the SRR, but performance was improved with low recruitment variability and contrast in fishing pressure. Some bias in estimation in at least one of the SRR parameters existed in nearly all OM-EM combinations. Because bias in terminal SSB and retrospective patterns were indifferent to whether or not the SRR was estimated, and convergence was slightly better without the SRR, a sensible default would be to fit models without an assumed SRR.

@@

\begin{landscape}
\begin{figure}
\begin{center}
\includegraphics{term_SSB_bias_plots}
\end{center}
\caption{Median relative error of terminal year SSB for estimating models fitted to data sets simulated with alternative process error structures: R and R+S (A), R+Sel (B), R+M (C), or R+q (D). Circled values indicate results where the EM process error structure matches that of the operating model and vertical lines represent 95\% confidence intervals.}\label{SSB_rel_error}
\end{figure}
\end{landscape}
